Deciding Bitvector Termination with Program Synthesis

### 2014-06-25 papers/termination/paper.tex
revised intro+related work

When surveying the area of program termination chronologically, we observe an initial domination of  monolithic approaches based on a single measure shown to decrease
over all program paths %(syntactic characterisation of loops) 
\cite{DBLP:conf/vmcai/P04,DBLP:conf/cav/BradleyMS05}, followed by 
more recent techniques moving towards termination arguments based on Ramsey's theorem \cite{DBLP:conf/lpe/CodishG03,DBLP:conf/lics/PodelskiR04,DBLP:conf/pldi/CookPR06}.
The latter approach aims to find a set of local termination conditions that decrease as execution follows through the loop. %(semantic characterisation of loops).
The main justification for this paradigm shift lies in the simplicity of the local termination measures when compared to the global ones, e.g.
there are cases in which proofs based on local measures involve applying linear functions while corresponding global
measures involve nonlinear functions or lexicographic orders. 
This explain why approaches based on Ramsey's theorem are restricted to searching for linear termination arguments.

@@

We propose a general framework that %does not assume the existence of linear ranking functions and can 
uniformly computes
\emph{lexicographic/non-lexicographic} \emph{linear/non-linear} 
ranking functions supported by inductive \emph{linear/non-linear} invariants for loops with 
\emph{linear/non-linear} guards and transitions over bit-vectors and floats.
%Our technique is {\emph complete} and can handle programs with non-linear operations, e.g. logical and.
%--In our design, the termination problem becomes as hard as finding ranking functions, rather than as hard as
%checking the validity of a termination argument.

For this purpose, we phrase the termination problem as a second order satisfaction problem and
we identify a fragment of second-order logic that is expressive enough to capture our formulation %several variations of the halting problem 
(Section~\ref{sec:second.order}), which we call \emph{the synthesis fragment}. We propose a method 
for checking the satisfiability of a formula in the synthesis fragment as program synthesis (Section~\ref{sec:synthesis}). 
Our method is theoretically complete, i.e. if a pair of a ranking function and a supporting invariant 
exist, it will be found. With respect to its performance, we show that its runtime is dominated by the size of 
the shortest termination proof.
%While investigating the search space and the distribution of solutions (Section~\ref{}), we identify factors 
%influencing our method's performance and
%find that our technique  
This means that our procedure is not directly dependent on the size of the analysed program or its number of variables, but on the Kolmogorov
complexity of its minimal termination argument. We argue that it is natural to expect programmers to write programs that have relatively short termination 
proofs as this will make them understandable.

%% Our approach to proving termination has two distinct features over previous work. 
%% First,
%% Second, our runtime is dependant on the KC of the shortest termination proof rather than the structure of the loop, number of variables, number of lassos...

%% We show that the probability of
%% a ranking function to have a low Kolmogorov complexity is higher than the probability to be linear (see Section~?).
%% Moreover, we find that our technique does actually gain from the generality of the termination argument. 
%% After investigating its theoretical bounds, we experimentally show that our technique performs well in practice.

@@

@@

@@

Proving program termination implies the simultaneous search for a termination argument and a supporting invariant.
For this purpose, several tools alternate 
between calls to a safety prover and calls to a ranking function synthesis tool \cite{}. 
In order to share more information about the state of the termination proofs between the two tools, 
Brockschmidt et al. use a single representation of the state of the termination proof search that both tools operate over, 
resulting in performance improvements \cite{DBLP:conf/cav/BrockschmidtCF13}.
In our setting, both the ranking function and the supporting invariant are iteratively constructed through a single refinement loop.

@@

\todo{add smth about synthesis}

>>>>

When surveying the area of program termination chronologically, we observe an initial domination of  monolithic approaches based on a single measure shown to decrease
over all program paths %(syntactic characterisation of loops) 
\cite{DBLP:conf/vmcai/P04,DBLP:conf/cav/BradleyMS05}, followed by 
more recent techniques moving towards termination arguments based on Ramsey's theorem \cite{DBLP:conf/lpe/CodishG03,DBLP:conf/lics/PodelskiR04,DBLP:conf/pldi/CookPR06}.
The latter approach aims to find a set of local termination conditions that decrease as execution follows through the loop. %(semantic characterisation of loops).
The main justification for this paradigm shift lies in the simplicity of the local termination measures when compared to the global ones, e.g.
there are cases in which proofs based on local measures involve applying linear functions while corresponding global
measures involve nonlinear functions or lexicographic orders. 
This explains why approaches based on Ramsey's theorem are restricted to searching for linear termination arguments.

@@

We propose a general framework that %does not assume the existence of linear ranking functions and can 
uniformly computes
\emph{lexicographic/non-lexicographic} \emph{linear/non-linear} 
ranking functions supported by inductive \emph{linear/non-linear} invariants for loops with 
\emph{linear/non-linear} guards and transitions over bit-vectors and floats.
%Our technique is {\emph complete} and can handle programs with non-linear operations, e.g. logical and.
%--In our design, the termination problem becomes as hard as finding ranking functions, rather than as hard as
%checking the validity of a termination argument. 
We combine the generation of ranking functions with the generation of invariants such that the necessary 
supporting invariants and the ranking function are discovered at the same time.

For this purpose, we phrase the termination problem as a second order satisfaction problem and
we identify a fragment of second-order logic that is expressive enough to capture our formulation, 
which we call \emph{the synthesis fragment}. Subsequently, we propose a method 
for checking the satisfiability of a formula in the synthesis fragment as program synthesis (Section~\ref{sec:synthesis}). 
Our method is theoretically complete, i.e. if a pair of a ranking function and a supporting invariant 
exist, it will be found.

With respect to the performance of our technique, we show that its runtime is dominated by the size of 
the shortest termination proof.
%While investigating the search space and the distribution of solutions (Section~\ref{}), we identify factors 
%influencing our method's performance and
%find that our technique  
This means that our procedure is not directly dependent on the structure of the analysed loop, e.g. number of variable, number of lassos, 
but on the Kolmogorov complexity of its minimal termination argument. 
\emph{In our design, proving a program terminates becomes as hard as finding its minimal temination argument}.
We argue that the length of a program's termination proof is indicative of how easy to understand the program is, and thus 
programmers tend to write programs that have relatively short termination proofs. 
We empirically show that this claim holds in practice, causing our technique to perform well.

%% Our approach to proving termination has two distinct features over previous work. 
%% First, it is \emph{general}, i.e. does not make any assumption about the form of the termination argument.   
%% Second, it correctly handles bit-vectors and floats without abstracting them to integers and reals, respectively.

@@

\item Our proposed termination proving procedure can also prove non-termination. Alternating termination and non-termination?

@@

Limitations: No recursion.

@@

Proving program termination implies the simultaneous search for a termination argument and a supporting invariant.
For this purpose, several tools alternate 
between calls to a safety prover and calls to a ranking function synthesis tool \cite{}. 
In order to share more information about the state of the termination proofs between the two tools, 
Brockschmidt et al. use a single representation of the state of the termination proof search that both tools operate over, 
resulting in performance improvements \cite{DBLP:conf/cav/BrockschmidtCF13}.
In \cite{DBLP:conf/cav/BradleyMS05}, Bradley et al. combine the generation of ranking functions with the generation of invariants to
form one constraint solving problem such that the necessary supporting invariants for the ranking function are discovered on demand.
In our setting, both the ranking function and the supporting invariant are iteratively constructed through the same refinement loop.

@@

\todo{add non-termination}
